Hilbert space methods to approximate Gaussian processes using Stan

### 2019-04-27 Paper/manuscript/manuscript_05.tex
fix a couple of figure references

Next, we will focus on analyzing the interaction effects between these $m$ and $c$ on the performance of the approximation. The lengthscale and marginal variance will no longer be fixed but rather estimated in both regular GP and HSGP models. Figure \ref{fig4_Post_&_Cov_part1} shows the functional posterior predictions and the covariance function obtained after fitting the data, for varying $m$ and $c$. Figure \ref{fig5_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models, computed against the regular GP model. Figure \ref{fig6_lscale_vs_J} shows the estimated lengthscale and marginal variance for the regular GP model and the HSGP models. Looking at the RMSEs in Figure \ref{fig5_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computations would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the choice of $m = 10$ and $c = 1.5$ could still be an accurate enough choice. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure \ref{fig4_Post_&_Cov_part1}. From these results, some general conclusions may be drawn:

@@

Additionally, there is a clear relation of the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figure \ref{fig3_lscale_vs_J_vs_c} depicts how these three factors interact with each other in relation to 
% the performance of the HSGP approximation.
a close approximation of the HSGP model, in the case of a GP with square exponential covariance function and a single input dimension. (Gabi: Could it be good to build and include the same Figure for a Matern($\frac{3}{2}$)?). More precisely, for a given GP model (with a square exponential covariance function) with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig3_lscale_vs_J_vs_c} shows the minimum $m$ required to achieve a close approximation in terms of satisfying Equation (\ref{diff_covs}). We have considered an approximation to be a close enough when the difference between densities of the approximate covariance function and the exact covariance function, $\varepsilon$ in Equation (\ref{diff_covs}), is below 1$\%$ of the density of the exact covariance function,
%
\begin{eqnarray*}
 \frac{\varepsilon}{\int k(\tau) \, d\tau} < 0.01.
\end{eqnarray*}

@@

As stated above, Figure \ref{fig3_lscale_vs_J_vs_c} provides the minimum lengthscale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the lenghscale $\ell$ controls the "wigglyness" of
the functional relationship, it strongly influences the difficulty of obtaining accurate
inference about the function from the data. Basically, if the lengthscale estimate is accurate, 
we can expect the HSGP approximation to be accurate as well. Thus, having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum lengthscale provided in Figure \ref{fig3_lscale_vs_J_vs_c}. If $\hat{\ell}$ exceeds this recommended minimum lengthscale, the approximation should be close enough. If, however, it does not exceed it, the approximation may be inaccurate and $m$ should be increased or $c$ decreased (Paul: decreasing $c$ seems dangerous to me if predictions for out of sample data are desired. Shell we mention this? Gabi: We assume that predicting points are already within $S$. Then, a valid $c$ is more related to the lengthscale than to the predicting points. In fact, valid $c$s are restricted in function of the lengthscale: A given lengthscale has a minimum $c$ under which a close approximation will never be achieved regarless $m$ (We stated this in line 502 and also in line 462). For example, a given $c$ can not work for certain lengthscale but for a larger one). We may also use this diagnostic in a iterative procedure.
Starting from some initial guess of $\ell$, we can choose initial values for $m$ and $c$ and fit an HSGP model, then check the approximation accuracy, and, if not accurate enough because the estimated $\hat{\ell}$ is below the minimum lengthscale provided by Figure \ref{fig3_lscale_vs_J_vs_c}, repeat the process while increasing $m$ or decreasing $c$. Note that, as commented before, $c$ can not be decreased as much as desired because it is restricted to the lengthscale.
%replacing $\ell$ by the newly obtained estimate $\hat{\ell}$.

@@

If we look back to the conclusions drawn from Figures \ref{fig5_MSE_vs_J} and \ref{fig6_lscale_vs_J}, where $m = 10$ basis functions and a boundary factor of $c = 1.5$ were enough to closely approximate a function with $\ell = 0.3$, we can recognize that these conclusions also matches those obtained from Figure \ref{fig3_lscale_vs_J_vs_c}.

@@

Figure \ref{fig3_lscale_vs_J_vs_c} was build for a GP with a unidimensional covariance function, and results in a surface depending of three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending of four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which can not be directly represented. However, as an approximation we can use the unidimensional GP conclusions in Figure \ref{fig3_lscale_vs_J_vs_c} to analyze individually the different dimensions of a multidimensional GP model.

@@

In this example, we make a comparison of the lengthscale estimates
obtained from the regular GP and HSGP models. We also have a look at those recommended minimum lengthscales provided by Figure \ref{fig3_lscale_vs_J_vs_c}.

@@

Figure \ref{fig7_varing_lscale} shows the posterior predictions of the regular GP and HSGP models fitted to those datasets. The Lengthscale estimates as obtained by regular GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models against the regular GP models as a function of the lengthscale and number of basis functions $m$.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.4, trim = 10mm 95mm 5mm 30mm, clip]{fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part9.pdf}}\\
\vspace{-3mm}
\caption{Posterior predictive means of both regular GP and HSGP models, fitted over different datasets simulated from square exponential GP models with different characteristic lengthscales ($lscale$) and same marginal variance ($\alpha$) as the data-generative function ($True function$).}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

@@

The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the minimum lengthscales that can be closely approximate in every case, according to Figure \ref{fig3_lscale_vs_J_vs_c}. It can be appreciated a complete agreement between those recommended minimum lengthscales that can be properly fitted in every case following Figure \ref{fig3_lscale_vs_J_vs_c} and the real approximate accuracy measured by the RMSE in Figure \ref{fig9_MSE_varing_lscale}.

>>>>

Next, we will focus on analyzing the interaction effects between these $m$ and $c$ on the performance of the approximation. The lengthscale and marginal variance will no longer be fixed but rather estimated in both regular GP and HSGP models. Figure \ref{fig3_Post_part1} shows the functional posterior predictions and the covariance function obtained after fitting the data, for varying $m$ and $c$. Figure \ref{fig4_MSE_vs_J} shows the root mean square error (RMSE) of the HSGP models, computed against the regular GP model. Figure \ref{fig5_lscale_vs_J} shows the estimated lengthscale and marginal variance for the regular GP model and the HSGP models. Looking at the RMSEs in Figure \ref{fig4_MSE_vs_J}, we can conclude that the optimal choice in terms of precision and computations would be $m = 15$ basis functions and a boundary factor between $c = 1.5$ and $c = 2.5$. Further, the choice of $m = 10$ and $c = 1.5$ could still be an accurate enough choice. We may also come to the same conclusion by looking at the posterior predictions and covariance function plots in Figure \ref{fig3_Post_part1}. From these results, some general conclusions may be drawn:

@@

Additionally, there is a clear relation of the number of basis functions $m$ and the boundary factor $c$ with the lengthscale $\ell$ of the approximated function. Figure \ref{fig6_lscale_vs_J_vs_c} depicts how these three factors interact with each other in relation to 
% the performance of the HSGP approximation.
a close approximation of the HSGP model, in the case of a GP with square exponential covariance function and a single input dimension. (Gabi: Could it be good to build and include the same Figure for a Matern($\frac{3}{2}$)?). More precisely, for a given GP model (with a square exponential covariance function) with lengthscale $\ell$ and given a boundary factor $c$, Figure \ref{fig6_lscale_vs_J_vs_c} shows the minimum $m$ required to achieve a close approximation in terms of satisfying Equation (\ref{diff_covs}). We have considered an approximation to be a close enough when the difference between densities of the approximate covariance function and the exact covariance function, $\varepsilon$ in Equation (\ref{diff_covs}), is below 1$\%$ of the density of the exact covariance function,
%
\begin{eqnarray*}
 \frac{\varepsilon}{\int k(\tau) \, d\tau} < 0.01.
\end{eqnarray*}

@@

As stated above, Figure \ref{fig6_lscale_vs_J_vs_c} provides the minimum lengthscale that can be closely approximated given $m$ and $c$. This information serves as a powerful diagnostic tool in determining if the obtained accuracy is acceptable. As the lenghscale $\ell$ controls the "wigglyness" of
the functional relationship, it strongly influences the difficulty of obtaining accurate
inference about the function from the data. Basically, if the lengthscale estimate is accurate, 
we can expect the HSGP approximation to be accurate as well. Thus, having obtained an estimate $\hat{\ell}$ of $\ell$ from the HSGP model based on prespecified $m$ and $c$, we can check whether or not $\hat{\ell}$ exceeds the minimum lengthscale provided in Figure \ref{fig6_lscale_vs_J_vs_c}. If $\hat{\ell}$ exceeds this recommended minimum lengthscale, the approximation should be close enough. If, however, it does not exceed it, the approximation may be inaccurate and $m$ should be increased or $c$ decreased (Paul: decreasing $c$ seems dangerous to me if predictions for out of sample data are desired. Shell we mention this? Gabi: We assume that predicting points are already within $S$. Then, a valid $c$ is more related to the lengthscale than to the predicting points. In fact, valid $c$s are restricted in function of the lengthscale: A given lengthscale has a minimum $c$ under which a close approximation will never be achieved regarless $m$ (We stated this in line 502 and also in line 462). For example, a given $c$ can not work for certain lengthscale but for a larger one). We may also use this diagnostic in a iterative procedure.
Starting from some initial guess of $\ell$, we can choose initial values for $m$ and $c$ and fit an HSGP model, then check the approximation accuracy, and, if not accurate enough because the estimated $\hat{\ell}$ is below the minimum lengthscale provided by Figure \ref{fig6_lscale_vs_J_vs_c}, repeat the process while increasing $m$ or decreasing $c$. Note that, as commented before, $c$ can not be decreased as much as desired because it is restricted to the lengthscale.
%replacing $\ell$ by the newly obtained estimate $\hat{\ell}$.

@@

If we look back to the conclusions drawn from Figures \ref{fig4_MSE_vs_J} and \ref{fig5_lscale_vs_J}, where $m = 10$ basis functions and a boundary factor of $c = 1.5$ were enough to closely approximate a function with $\ell = 0.3$, we can recognize that these conclusions also matches those obtained from Figure \ref{fig6_lscale_vs_J_vs_c}.

@@

Figure \ref{fig6_lscale_vs_J_vs_c} was build for a GP with a unidimensional covariance function, and results in a surface depending of three variables, $m$, $c$ and $\ell$. An equivalent figure for a GP model with a two-dimensional covariance function would result in a surface depending of four variables, $m$, $c$, $\ell_1$ and $\ell_2$, which can not be directly represented. However, as an approximation we can use the unidimensional GP conclusions in Figure \ref{fig6_lscale_vs_J_vs_c} to analyze individually the different dimensions of a multidimensional GP model.

@@

In this example, we make a comparison of the lengthscale estimates
obtained from the regular GP and HSGP models. We also have a look at those recommended minimum lengthscales provided by Figure \ref{fig6_lscale_vs_J_vs_c}.

@@

Figure \ref{fig7_posterior_varing_lscale_part1} shows the posterior predictions of the regular GP and HSGP models fitted to those datasets. The Lengthscale estimates as obtained by regular GP and HSGP models are depicted in Figure \ref{fig8_Tlscale_vs_Elscale}. Further, Figure \ref{fig9_MSE_varing_lscale} shows the root mean square error (RMSE) of the HSGP models against the regular GP models as a function of the lengthscale and number of basis functions $m$.

\begin{figure}
\centering
\subfigure{\includegraphics[scale=0.4, trim = 10mm 95mm 5mm 30mm, clip]{fig7_legend.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part1.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part2.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part3.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part4.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part5.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 31mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part6.pdf}}\\
\vspace{-3mm}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part7.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part8.pdf}}
\subfigure{\includegraphics[scale=0.29, trim = 10mm 10mm 10mm 19mm, clip]{fig7_posterior_varing_lscale_part9.pdf}}\\
\vspace{-3mm}
\caption{Posterior predictive means of both regular GP and HSGP models, fitted over different datasets simulated from square exponential GP models with different characteristic lengthscales ($lscale$) and same marginal variance ($\alpha$) as the data-generative function (\textit{True function}).}
  \label{fig7_posterior_varing_lscale_part1}
\end{figure}

@@

The dashed black line in Figure \ref{fig8_Tlscale_vs_Elscale} represents the minimum lengthscales that can be closely approximate in every case, according to Figure \ref{fig6_lscale_vs_J_vs_c}. It can be appreciated a complete agreement between those recommended minimum lengthscales that can be properly fitted in every case following Figure \ref{fig6_lscale_vs_J_vs_c} and the real approximate accuracy measured by the RMSE in Figure \ref{fig9_MSE_varing_lscale}.
